---
title: "Frequently Asked Questions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Frequently Asked Questions

## The X axis should be "1-Specificity" instead of "Specificity"!

This is not a bug! 

Most statistical tools cannot plot specificity from 1 to 0, so instead they plot
1-specificity from 0 to 1. This results in exactly the same plot, just with a
different labelling of the axis.

R doesn't have this limitation, and `pROC` plots specificity from 1 to 0, saving
you from doing a few subtractions when you look at the plot. If you are using
pROC in S+, you only have the choice to plot 1-specificity. 

If you are concerned that your reviewers will be confused and really wish to
have the "old-style" 1-Specificity from 0 to 1, or if you plan to change the
axis label to "False Positive Rate" (which is really 1-specificity), you can use
the `legacy.axis` argument to the plot function. See `?plot.roc` for more
details.

## I have a GLM/SVM/PLS/<INSERT YOUR FAVORITE MODEL HERE> model, how can I analyze it with pROC?

You need to use the `predict()` function to get predictions that can be passed
to pROC (as the `predictor` argument).

Specifically you need to get a single vector of numeric predictions. In some
cases you may have to add pass `type="probabilities"` or something similar to
the `predict` function (to get probabilities instead of the binary response),
sometimes you may to have select one of the column of the returned matrix (if
`predict` returns a matrix of class probabilities). Read the documentation of
the package you're using and look at your results to find out what to do
exactly.

## Can I create a ROC curve with two predictors? Can I adjust fixed / random effects in pROC?

Not directly, but you can! You need to use a modeling function (such as `glm()`)
or package (such as [lme4](http://cran.r-project.org/web/packages/lme4/)); fit
and check the model, then use the `predict()` function to get predictions that
can be passed to pROC (as the `predictor` argument).

## When I resample or randomize my dataset I get an biased AUC > 0.5. How come?

This is because pROC tries to auto-detect the direction of the decision due to
the default value of `direction="auto"`. The decision is based on the median of
the groups instead of the AUC directly, so you the effect may not be 1:1, but
you should expect to obtain AUC > 0.5 on average.

Whenever you are resampling with pROC you need to explicitly pass an explicit
value to `direction`, either `direction="<"` or `direction=">"`. I also
recommend taking a minute to set the `levels` (value of the `response` for
controls and cases, respectively) argument explicitly too.

A future version of pROC will be more verbose and print a message when auto-detecting the direction.

## Can I test if a single ROC curve is significantly different from 0.5?

This is equivalent to asking whether the median values of cases and controls are
significantly different. Therefore you can easily perform a [Wilcoxon Rank Sum
Test](https://www.rdocumentation.org/packages/stats/topics/wilcox.test).

## My dataset is huge. Can I use pROC?

A large effort has been made to render pROC more and more efficient with large
ROC curves. An algorithm with complexity of O(n) has been rolled out with pROC
1.6 and improved over time. Improvements to the DeLong test and the `coords`
functions have been implemented too.

If you feel that a function is abnormally slow, especially in comparison to
other packages, feel free to [open a bug
report](https://github.com/xrobin/pROC/issues/new?template=Bug_report.md).

## Can I compute the confidence interval of the best threshold of the curve (not its SE/SP)?

Yes you can since pROC 1.6 and its `ci.coords` function. For instance:

```{r, eval = FALSE}
ci.coords(aSAH$outcome, aSAH$s100b, x="best", input = "threshold",
          ret="threshold")
# 95% CI (2000 stratified bootstrap replicates):
#            2.5%   50%  97.5%
# threshold 0.115 0.205 0.4854
```

There is 95% chance that the optimal threshold lies between 0.115 and 0.4854.

## The functions are giving me weird error messages I don't understand

Several packages on CRAN provide alternative `roc` or `auc` functions. These
packages can interfere with pROC, especially if they are loaded later in the
session (and hence appear earlier in the search path).

For instance, here are a few messages you may see if you have the `AUC` package loaded:

```{r, eval = FALSE}
roc(outcome ~ ndka, data = aSAH)
# Not enough distinct predictions to compute area under the ROC curve.
# 
# Error in roc(outcome ~ ndka, data = aSAH) : unused argument (data = aSAH)

If that happens, you should unload the package by typing `detach("package:AUC")`.
To find out where the function is coming from, simply type its name on the R
prompt:

```r
roc
```

The line at the bottom indicates that the function comes from the AUC package.

In addition, you may have defined a `roc` function yourself. In this case,
`rm(roc)` will solve the problem.

Alternatively you can refer to the pROC version of the function specifically
through their namespace:

```r
pROC::auc(pROC::roc(aSAH$outcome, aSAH$ndka))
```

Here is a list of packages providing `roc`, `auc` and `ci` functions which will
hide pROC's functions if the package is loaded after pROC:

| Package | roc | auc | ci |
|---------|-----|-----|-----|
| analogue | [Generic](http://finzi.psych.upenn.edu/R/library/analogue/html/roc.html) | | |
| asbio | | [Not Generic](http://finzi.psych.upenn.edu/R/library/asbio/html/auc.html) | |
| AUC | [Not Generic](http://finzi.psych.upenn.edu/R/library/AUC/html/roc.html) | [Not Generic](http://finzi.psych.upenn.edu/R/library/AUC/html/auc.html) | |
| aucm | [NA](http://finzi.psych.upenn.edu/R/library/aucm/html/roc.html) | [NA](http://finzi.psych.upenn.edu/R/library/aucm/html/auc.html) | |
| cutpointr | [Not Generic](http://finzi.psych.upenn.edu/R/library/cutpointr/html/roc.html) | | |
| distillery | | | [Generic](http://finzi.psych.upenn.edu/R/library/distillery/html/ci.html) |
| enrichvs | | [Not Generic](http://finzi.psych.upenn.edu/R/library/enrichvs/html/auc.html) | |
| epiDisplay | [NA](http://finzi.psych.upenn.edu/R/library/epiDisplay/html/roc.html) | | [Generic](http://finzi.psych.upenn.edu/R/library/epiDisplay/html/ci.html) |
| esvis | | [Not Generic](http://finzi.psych.upenn.edu/R/library/esvis/html/auc.html) | |
| FFTrees | | [Not Generic](http://finzi.psych.upenn.edu/R/library/FFTrees/html/auc.html) | |
| flux | | [Not Generic](http://finzi.psych.upenn.edu/R/library/flux/html/auc.html) | |
| fmsb | [Not Generic](http://finzi.psych.upenn.edu/R/library/fmsb/html/roc.html) | | |
| gmodels | | | [Generic](http://finzi.psych.upenn.edu/R/library/gmodels/html/ci.html) |
| grpreg | | [NA](http://finzi.psych.upenn.edu/R/library/grpreg/html/auc.html) | |
| hutils | | [Not Generic](http://finzi.psych.upenn.edu/R/library/hutils/html/auc.html) | |
| kulife | | [Not Generic](http://finzi.psych.upenn.edu/R/library/kulife/html/auc.html) | |
| longROC | [Not Generic](http://finzi.psych.upenn.edu/R/library/longROC/html/roc.html) | [Not Generic](http://finzi.psych.upenn.edu/R/library/longROC/html/auc.html) | |
| MAMSE | [Not Generic](http://finzi.psych.upenn.edu/R/library/MAMSE/html/roc.html) | | |
| MESS | | [Not Generic](http://finzi.psych.upenn.edu/R/library/MESS/html/auc.html) | |
| Metrics | | [Not Generic](http://finzi.psych.upenn.edu/R/library/Metrics/html/auc.html) | |
| MXM | | [Not Generic](http://finzi.psych.upenn.edu/R/library/MXM/html/auc.html) | |
| ncvreg | | [NA](http://finzi.psych.upenn.edu/R/library/ncvreg/html/auc.html) | |
| NEArender | [Not Generic](http://finzi.psych.upenn.edu/R/library/NEArender/html/roc.html) | | |
| PK | | [Not Generic](http://finzi.psych.upenn.edu/R/library/PK/html/auc.html) | |
| precrec | | [Generic](http://finzi.psych.upenn.edu/R/library/precrec/html/auc.html) | |
| PresenceAbsence | | [Not Generic](http://finzi.psych.upenn.edu/R/library/PresenceAbsence/html/auc.html) | |
| radiant.model | | [Not Generic](http://finzi.psych.upenn.edu/R/library/radiant.model/html/auc.html) | |
| RMediation | | | [Not Generic](http://finzi.psych.upenn.edu/R/library/RMediation/html/ci.html) |
| sdm | [Generic](http://finzi.psych.upenn.edu/R/library/sdm/html/roc.html) | | |
| SDMTools | | [Not Generic](http://finzi.psych.upenn.edu/R/library/SDMTools/html/auc.html) | |
| spatstat | | [Generic](http://finzi.psych.upenn.edu/R/library/spatstat/html/auc.html) | |
| StatMeasures | | [Not Generic](http://finzi.psych.upenn.edu/R/library/StatMeasures/html/auc.html) | |
| survMisc | | | [Generic](http://finzi.psych.upenn.edu/R/library/survMisc/html/ci.html) |

To get an updated list, run `inst/extra/sos_clashes.R`. It may be needed to
start R with `export R_MAX_NUM_DLLS=1000` to be able to run the script.
