---
title: "Plotting ROC curves with ggplot2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting ROC curves with ggplot2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  dpi = 120
)
```

This vignette recreates the six classic `plot.roc` figures from the
[pROC screenshots gallery](https://xrobin.github.io/pROC/screenshots.html)
using `ggroc` and the ggplot2 layers introduced in pROC 1.20.0.

Bootstrap confidence intervals use `boot.n = 100` here so the vignette
stays fast. The gallery uses the default `boot.n = 2000`.

```{r setup}
library(pROC)
library(ggplot2)
data(aSAH)

rocobj <- roc(aSAH$outcome, aSAH$s100b, percent = TRUE)
roc_ndka <- roc(aSAH$outcome, aSAH$ndka, percent = TRUE)

set.seed(42)
rocobj_ci <- roc(aSAH$outcome, aSAH$s100b, percent = TRUE, ci = TRUE, boot.n = 100)
ci_se <- ci.se(rocobj_ci, specificities = seq(0, 100, 5), boot.n = 100)
ci_sp <- ci.sp(rocobj_ci, sensitivities = seq(0, 100, 5), boot.n = 100)
ci_th <- ci.thresholds(rocobj_ci, thresholds = "best", boot.n = 100)
```

```{r helpers}
theme_ggroc_screenshot <- function(roc, legend.position = "none") {
  list(
    geom_roc_identity(roc),
    theme_bw(),
    theme(
      legend.position = legend.position,
      legend.justification = if (identical(legend.position, c(1, 0))) c(1, 0) else "center",
      legend.background = element_rect(fill = "white", colour = NA),
      panel.grid.minor = element_blank()
    )
  )
}
```

## 1. Partial AUC (pAUC)

```{r screenshot-1-classic}
plot.roc(aSAH$outcome, aSAH$s100b,
  percent = TRUE,
  partial.auc = c(100, 90), partial.auc.correct = TRUE,
  print.auc = TRUE,
  print.auc.pattern = "Corrected pAUC (100-90%% SP):\n%.1f%%",
  print.auc.col = "#1c61b6",
  auc.polygon = TRUE, auc.polygon.col = "#1c61b6",
  max.auc.polygon = TRUE, max.auc.polygon.col = "#1c61b622",
  main = "Partial AUC (pAUC)"
)
plot.roc(aSAH$outcome, aSAH$s100b,
  percent = TRUE, add = TRUE, type = "n",
  partial.auc = c(100, 90), partial.auc.correct = TRUE,
  partial.auc.focus = "se",
  print.auc = TRUE,
  print.auc.pattern = "Corrected pAUC (100-90%% SE):\n%.1f%%",
  print.auc.col = "#008600",
  print.auc.y = 40,
  auc.polygon = TRUE, auc.polygon.col = "#008600",
  max.auc.polygon = TRUE, max.auc.polygon.col = "#00860022"
)
```

```{r screenshot-1-ggroc}
auc_sp <- auc(rocobj, partial.auc = c(100, 90), partial.auc.correct = TRUE)
auc_se <- auc(rocobj,
  partial.auc = c(100, 90), partial.auc.correct = TRUE,
  partial.auc.focus = "se"
)

ggroc(rocobj) +
  geom_polygon_max_auc(auc_se, fill = "#008600", alpha = 0x22 / 255) +
  geom_polygon_max_auc(auc_sp, fill = "#1c61b6", alpha = 0x22 / 255) +
  geom_polygon_auc(auc_se, fill = "#008600", colour = NA) +
  geom_polygon_auc(auc_sp, fill = "#1c61b6", colour = NA) +
  geom_text_auc(auc_sp,
    pattern = "Corrected pAUC (100-90%% SP):\n%.1f%%",
    colour = "#1c61b6"
  ) +
  geom_text_auc(auc_se,
    y = 40,
    pattern = "Corrected pAUC (100-90%% SE):\n%.1f%%",
    colour = "#008600"
  ) +
  ggtitle("Partial AUC (pAUC)") +
  theme_ggroc_screenshot(rocobj)
```

## 2. Confidence intervals

```{r screenshot-2-classic}
plot(rocobj_ci,
  main = "Confidence intervals", percent = TRUE,
  print.auc = TRUE
)
plot(ci_se, type = "shape", col = "#1c61b6AA")
plot(ci_th)
```

```{r screenshot-2-ggroc}
ggroc(rocobj_ci) +
  geom_polygon_ci(ci_se, fill = "#1c61b6", alpha = 0xAA / 255) +
  geom_errorbar_ci(ci_th) +
  geom_text_auc(rocobj_ci) +
  ggtitle("Confidence intervals") +
  theme_ggroc_screenshot(rocobj)
```

## 3. Smoothing

```{r screenshot-3-classic}
plot.roc(aSAH$outcome, aSAH$s100b, percent = TRUE, main = "Smoothing")
lines(smooth(rocobj), col = "#1c61b6")
lines(smooth(rocobj, method = "density"), col = "#008600")
lines(smooth(rocobj, method = "fitdistr", density = "lognormal"), col = "#840000")
legend("bottomright",
  legend = c("Empirical", "Binormal", "Density", "Fitdistr\n(Log-normal)"),
  col = c("black", "#1c61b6", "#008600", "#840000"), lwd = 2
)
```

```{r screenshot-3-ggroc}
smoothed <- list(
  Empirical = rocobj,
  Binormal = smooth(rocobj),
  Density = smooth(rocobj, method = "density"),
  `Fitdistr (Log-normal)` = smooth(rocobj, method = "fitdistr", density = "lognormal")
)

ggroc(smoothed) +
  geom_roc_identity(rocobj) +
  scale_colour_manual(values = c(
    Empirical = "black",
    Binormal = "#1c61b6",
    Density = "#008600",
    `Fitdistr (Log-normal)` = "#840000"
  )) +
  ggtitle("Smoothing") +
  theme_bw() +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", colour = NA),
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## 4. Confidence intervals of specificity / sensitivity

```{r screenshot-4-classic}
plot(rocobj,
  main = "Confidence intervals of specificity/sensitivity",
  percent = TRUE
)
plot(ci_se, type = "shape", col = "#1c61b6AA")
plot(ci_sp, type = "bars")
```

```{r screenshot-4-ggroc}
ggroc(rocobj) +
  geom_polygon_ci(ci_se, fill = "#1c61b6", alpha = 0xAA / 255) +
  geom_errorbar_ci(ci_sp) +
  ggtitle("Confidence intervals of specificity/sensitivity") +
  theme_ggroc_screenshot(rocobj)
```

## 5. Confidence interval of a threshold

```{r screenshot-5-classic}
plot(rocobj,
  main = "Confidence interval of a threshold", percent = TRUE,
  print.thres = "best"
)
plot(ci_th)
```

```{r screenshot-5-ggroc}
ggroc(rocobj) +
  geom_errorbar_ci(ci_th) +
  geom_roc_threshold(rocobj, thresholds = "best") +
  ggtitle("Confidence interval of a threshold") +
  theme_ggroc_screenshot(rocobj)
```

## 6. Comparison

```{r screenshot-6-classic}
rocobj1 <- plot.roc(aSAH$outcome, aSAH$s100b,
  main = "Statistical comparison", percent = TRUE, col = "#1c61b6"
)
rocobj2 <- lines.roc(aSAH$outcome, aSAH$ndka, percent = TRUE, col = "#008600")
testobj <- roc.test(rocobj1, rocobj2)
text(50, 50, labels = paste("p-value =", format.pval(testobj$p.value)), adj = c(0, .5))
legend("bottomright", legend = c("S100B", "NDKA"),
       col = c("#1c61b6", "#008600"), lwd = 2)
```

```{r screenshot-6-ggroc}
testobj <- roc.test(rocobj, roc_ndka)

ggroc(list(S100B = rocobj, NDKA = roc_ndka)) +
  geom_roc_identity(rocobj) +
  annotate("text", x = 50, y = 50,
           label = paste("p-value =", format.pval(testobj$p.value)),
           hjust = 0, vjust = 0.5) +
  scale_colour_manual(values = c(S100B = "#1c61b6", NDKA = "#008600")) +
  ggtitle("Statistical comparison") +
  theme_bw() +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", colour = NA),
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```
